Supplementary Information for: Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal

### 2020-02-13 writeup/si.tex
code

%NOTE the proof could be made simpler by taking $m_t$ to be a deterministic function of $x_t$, $m_{t-1}$, rather than talking about conditional independence
%
%We first show a lemma:
%
%\begin{lemma}
%For any positive integer $t$, the following inequality holds:
%\begin{equation}
%H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
%\end{equation}
%\end{lemma}
%
%\begin{proof}[Proof of the Lemma]
Because $m_t$ is determined by $(w_{1 \dots t-1}, m_1)$:
\begin{equation}
	m_t = M(m_{t-1}, w_{t-1}) = M(M(m_{t-2}, w_{t-2}), w_{t-1}) = M(M(M(m_{t-3}, w_{t-3}), w_{t-2}), w_{t-1}) = \dots
\end{equation}
the Data Processing inequality entails the following inequality for every positive integer $t$:
\begin{equation}
	H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
\end{equation}

%\end{proof}

%\begin{proof}[Proof of the Lemma]
%	By Bayes' Theorem
%\begin{align*}
%	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
%\end{align*}
%By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
%\begin{align*}
%p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
%\end{align*}
%So we have a Markov chain
%\begin{equation}
%(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
%\end{equation}
%Thus, by the Data Processing Inequality,
%\begin{equation}
%H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
%\end{equation}
%Finally, iteratively applying this inequality, we get:
%\begin{align*}
%H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
%\end{align*}
%\end{proof}
Plugging this inequality into Equation~\ref{eq:byStation} above:
\begin{align}\label{eq:plugged}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
& = \frac{1}{T} \left(I[w_{1\dots T}, w_{\leq 0}] - I[w_{1\dots T}, m_1]\right) 
\end{align}
The first term $I[w_{1\dots T}, w_{\leq 0}]$ can be rewritten in terms of $I_t$:
\begin{align}\label{eq:i-expanded}
I[w_{1\dots T}, w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
\end{align}
Therefore
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}, m_1]\right) 
\end{align*}
$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
&= \sum_{t > T} I_t
\end{align*}
Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
%\end{proof}

%Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.

@@

Here we show results linking memory and locality in production.

First, we consider a setting in which a speaker produces sentences with bounded memory, and analyze the deviation of the produced distribution from the actual distribution of the language.

The first setting does not account for the fact that language is produced aiming for some communicative goal.
We therefore now assume that the speaker has a communicative goal $G$ in mind.
This goal $G$ stays constant during production process for a sentence, and we count how much memory is needed in addition to the goal $G$.
We assume that there is a distribution of sentences expressing goals $G$:
\begin{equation}
	P(sentence|G)
\end{equation}
and assume that the speaker aims to match this distribution
\begin{equation}
\mathbb{E}_G[D_{KL}((language|G)||(produced|G))]
\end{equation}
We can analyze this model by adding conditioning w.r.t. $G$ throughout the analysis of the previous case.
Specifically, we need $I_t^G := I[X_t, X_0|X_1, \dots, X_{t-1}, G]$.

@@

@@

>>>>

%NOTE the proof could be made simpler by taking $m_t$ to be a deterministic function of $x_t$, $m_{t-1}$, rather than talking about conditional independence
%
%We first show a lemma:
%
%\begin{lemma}
%For any positive integer $t$, the following inequality holds:
%\begin{equation}
%H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
%\end{equation}
%\end{lemma}
%
%\begin{proof}[Proof of the Lemma]
Because $m_t$ is determined by $(w_{1 \dots t-1}, m_1)$:
\begin{equation}
	m_t = M(m_{t-1}, w_{t-1}) = M(M(m_{t-2}, w_{t-2}), w_{t-1}) = M(M(M(m_{t-3}, w_{t-3}), w_{t-2}), w_{t-1}) = \dots
\end{equation}
the Data Processing inequality entails the following inequality for every positive integer $t$:

	Plugging this inequality into Equation~\ref{eq:byStation} above:
\begin{align}\label{eq:plugged}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
& = \frac{1}{T} \left(I[w_{1\dots T}, w_{\leq 0}] - I[w_{1\dots T}, m_1]\right) 
\end{align}
The first term $I[w_{1\dots T}, w_{\leq 0}]$ can be rewritten in terms of $I_t$:
\begin{align}\label{eq:i-expanded}
I[w_{1\dots T}, w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
\end{align}
Therefore
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}, m_1]\right) 
\end{align*}
$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
&= \sum_{t > T} I_t
\end{align*}
Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
%\end{proof}

%Justifylinear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.

\subsection{For nondeterministic encoding functions}

We relax condition (X) to
\begin{equation}\label{eq:listener-markov-nondeterministic}
p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
\end{equation}
This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.

The one place in the proof where (X) plays a role is the proof of the inequality:
\begin{equation}
	H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
\end{equation}

We show that this inequality still holds under the relaxed condition (X):
\begin{proof}
	By Bayes' Theorem
\begin{align*}
	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
\end{align*}
By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
\begin{align*}
p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
\end{align*}
So we have a Markov chain
\begin{equation}
(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
\end{equation}
Thus, by the Data Processing Inequality,
\begin{equation}
H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
\end{equation}
Finally, iteratively applying this inequality, we get:
\begin{align*}
H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
\end{align*}
\end{proof}

@@

Here we show results linking memory and locality in production.
We show that results similar to our main theorem hold for the tradeoff between a speaker's memory and the accuracy with which they match the distribution of the language.

\paragraph{Speaker aims to match language distribution}
First, we consider a setting in which a speaker produces sentences with bounded memory, and analyze the deviation of the produced distribution from the actual distribution of the language.

We consider a speaker who maintains memory representations and incrementally produces based on these representations:
\begin{equation}
	p_{speaker}(x_t|X_{<t}) = p(x_t|m_t) \\
\end{equation}
We show a tradeoff between the memory capacity $\operatorname{H}[m_t]$ and the KL-divergence between the actual language statistics and the speaker's production distribution:
\begin{equation}
D_{KL}(P_{language}||P_{produced})  := \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log \frac{p(x_t|X_{<t})}{p_{speaker}(x_t|X_{<t})} \\
\end{equation}
\begin{thm}
If a speaker maintains memory
	\begin{equation}
		\operatorname{H}[m_t] \leq \sum_{i=1}^T tI_t
	\end{equation}
	then 
\begin{equation}
	D_{KL}(P_{language}||P_{produced}) \geq \sum_{t=T+1}^\infty I_t
\end{equation}
\end{thm}

While this bound only considers the production of a single word, it immediately entails a bound on the production accuracy for sequences:
\begin{equation}
	D_{KL}(P_{language}(X_1\dots X_t|X_{\leq 0})||P_{produced}(X_1\dots X_t|X_{\leq 0}))  = t \cdot D_{KL}(P_{language}(X_1|X_{\leq 0})||P_{produced}(X_1|X_{\leq 0}))
\end{equation}

\begin{proof}
First note
\begin{align}
D_{KL}(P_{language}||P_{produced}) & = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log \frac{p(x_t|X_{<t})}{p_{speaker}(x_t|X_{<t})} \\
& = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log \frac{p(x_t|X_{<t})}{p(x_t|M(X_{<t}))} \\
& = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log p(x_t|X_{<t}) - \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log p(x_t|M(X_{<t})) \\
	& = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log p(x_t|X_{<t}) + S_M(x_t|X_{<t})
\end{align}
In the last line, the first term is a constant independent of $M$.

Then the proof for the listener case transfers to this setting.
\end{proof}

TODO limitations of this as a model of production

\paragraph{Speaker aims to match, conditional on goal}
The first setting does not account for the fact that language is produced aiming for some communicative goal.
We therefore now assume that the speaker has a communicative goal $G$ in mind.
This goal $G$ stays constant during production process for a sentence, and we count how much memory is needed in addition to the goal $G$.
We assume that there is a distribution of sentences expressing goals $G$:
\begin{equation}
	P(sentence|G)
\end{equation}
and assume that the speaker aims to match this distribution
\begin{equation}
\mathbb{E}_G[D_{KL}((language|G)||(produced|G))]
\end{equation}
We can analyze this model by adding conditioning w.r.t. $G$ throughout the analysis of the previous case.
Specifically, we need
\begin{equation}
I_t^G := I[X_t, X_0|X_1, \dots, X_{t-1}, G]
\end{equation}

@@

\begin{thm}
If a speaker maintains memory
	\begin{equation}
		\operatorname{H}[m_t] \leq \sum_{i=1}^T tI_t^G
	\end{equation}
	then 
\begin{equation}
\E_G	D_{KL}(P_{language}(\cdot|G)||P_{produced}(\cdot|G)) \geq \sum_{t=T+1}^\infty I_t^G
\end{equation}
\end{thm}

\begin{proof}
This is entirely analogous to the previous proof.
\end{proof}

are there conditions under which this is close to $I_t$?

\paragraph{Pragmatic Speaker}

would need an assumption on the density of goals in the space of sequences.

Note 
\begin{equation}
	D_{KL}(P_{language}(x_{1\dots t})||P_{produced}(x_{1\dots t}))  := \sum_{x_{1\dots t}} p(x_t|X_{1\dots t}) \log \frac{p(x_t|X_{1\dots t})}{p_{speaker}(x_t|X_{1\dots t})} \geq t D_KL(x_t||...)
\end{equation}

$H[G|Produced] - H[G|Language]$

\section{Proof of Left-Right Invariance}

Here we show that the bound provided by our theorem is invariant under reversal of the process.
That is: Given a process $(X_t)_{t \in \mathbb{Z}}$, we define its reverse process $(Y_t)_{t \in \mathbb{Z}}$ by $Y_t := X_t$.
We claim that the theorem provides the same bounds for the memory-surprisal tradeoff curves.
To prove this, we note:
\begin{equation}
	I[X_t, X_0|X_{1\dots t-1}] = I[Y_{-t}, Y_0|Y_{1-t\dots -1}] = I[Y_0, Y_t|Y_{1\dots t-1}] = I[Y_t, Y_0|Y_{1\dots t-1}]
\end{equation}
The first step follows from the definition of $Y$. The second step follows from the fact that $X_t$, and thus also $Y_t$, is stationary, and thus adding $t$ to each index in the expression does not change the resulting value. The third step uses the fact that mutual information is symmetric.

@@

\section{Optimization}
We cannot optimize for AUC, because we cannot construct an unbiased gradient estimator
$AUC \propto \sum_{t} (\sum_{s \leq t} I_s) tI_t = \sum_{s \leq t}  tI_t I_s$

Softmax gradient update corresponds to adding $\alpha$ to the target logit and removing $\alpha$ from all other logits.
Equivalently, add a certain dynamic amount to the logcount of the target and the total logcount.

Counting corresponds to adding $1$ to the target probability.
